Peaks and valleys of prolatin-related gene expression during pigeon parental care stages

### 2019-09-07 docs/characterization_manuscript.tex
now with 5 figs

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit.

@@

Understanding the mechanisms underlying parental care are critical to
circumventing issues with parent-newborn bonding as well, where ultimate
explanations are obvious, but specific mechanisms remain elusive. The
rock dove (\emph{Columba livia}) is an ideal system to characterize
changes in genetic expression during parental care transitions because:
1) ample genomic resources are available (Gillespie et al.~2013),
including a complete annotated genome assembly {[}1{]} and methodology
concerning reproductive physiology and behavior (Dong et al.~2012); and
2) rock doves are prolific, year-round breeders that thrive in
captivity, making observation, manipulation, and sampling highly
feasible year-round.

Rock doves are socially monogamous and offer bi-parental care, making
inter- and intra-sexual comparisons possible. Birds offer two important
behavioral transition points into parental care: the incubation of eggs
and the caring for chicks. This produces two unique opportunities to
study how the brain transitions into two different suites of parental
care behaviors. Additionally, rock doves exhibit a parental care
strategy analogous to mammals in that they, too, `lactate' to feed their
young (Gillespie et al.~2011, 2012). This lactation, unlike simple
regurgitation of food, consists of the production and sloughing off of
skin cells inside the crop sac of females and males, creating a
protein-rich milk-like substance on which they rear their chicks. Many
functional similarities between rock dove and mammalian lactation exist
concerning the mediation of this event by the hormone prolactin (Dumont
1965). Additionally, like mammalian milk, rock dove milk delivers
essential immunoglobulins and nutritional benefits to young, aiding in
their immune function and development of microbiota {[}2{]}. Thus,
because rock doves incubate eggs and exhibit mammalian-like mediation
and function of lactation for young, they have the potential to serve as
a powerful theoretical bridge to understand the neurobiology of both
avian and mammalian transitions into parental care. Our working
hypothesis is that distinct changes in transcription occur in the brain
at the anticipation of, during, and in response to two different types
of parental care: incubation behavior and hatchling care.

@@

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}

We used EdgeR and DESeq2 to calculate statistical differnces in gene
expression between each transition. All subsequent analyses were
conducting using variance stabilized data from the DESeq2 model.

@@

Thosands of genes were found to be differentially expressed between the
non-parental control group and the building group (Table 1, Fig. 1).
Within the time course of parental care, fewer than 1000 genes change
there expression over the course of a few days, with the most pronounced
difference in toltal number of changes occuring in the female pituitary
between incubation days 9 and 17 (Table 1, Fig. 1).

\begin{table}[]
\begin{tabular}{llllllll}
\hline
\textbf{Method} & \textbf{Contrast} & \textbf{Hyp F} & \textbf{Hyp M} & \textbf{Pit F} & \textbf{Pit M} & \textbf{Gon F} & \textbf{Gon M} \\  \hline
EdgeR & Cont - Bldg & 14674 & 117 & 14593 & 144 & 14859 & 4 \\
EdgeR & Bldg - Lay & 14742 & 0 & 14593 & 0 & 14836 & 0 \\
EdgeR & Lay - Inc3 & 0 & 0 & 0 & 0 & 19 & 0 \\
EdgeR & Inc3 - Inc9 & 0 & 1 & 0 & 3 & 2 & 0 \\
EdgeR & Inc9 - Inc 17 & 0 & 2 & 57 & 37 & 9 & 0 \\
EdgeR & Inc17 - Hatch & 0 & 0 & 3 & 1 & 3 & 0 \\
EdgeR & Hatch - Nest5 & 1 & 0 & 3 & 0 & 0 & 0 \\
EdgeR & Nest5 - Nest9 & 0 & 0 & 0 & 0 & 24 & 0 \\  \hline
DESeq2 & Cont - Bldg & 3452 & 4499 & 3950 & 4680 & 4597 & 2135 \\
DESeq2 & Bldg - Lay & 0 & 0 & 101 & 0 & 104 & 1 \\
DESeq2 & Lay - Inc3 & 0 & 0 & 18 & 1 & 440 & 1 \\
DESeq2 & Inc3 - Inc9 & 1 & 0 & 0 & 1 & 8 & 0 \\
DESeq2 & Inc9 - Inc 17 & 1 & 0 & 640 & 186 & 1 & 0 \\
DESeq2 & Inc17 - Hatch & 2 & 0 & 3 & 18 & 1 & 0 \\
DESeq2 & Hatch - Nest5 & 166 & 0 & 340 & 3 & 1 & 0 \\
DESeq2 & Nest5 - Nest9 & 0 & 0 & 1 & 23 & 19 & 0 \\ \hline
\end{tabular}
\caption{Comparison of total DEGs determined by edgeR and DESeq2}
\label{tab:my-table}
\end{table}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}

\textbf{Fig 1. The magnitude of gene expression changes between each
parental transition}. Using DESeq 2, we identified 4-5K genes are
differentially expressed beteen control birds and their nest buliding
conspecifics in all tissues except the male gonad. 500 - 1000 genes are
differentially expressed in the female pituitary from mid-late
incubation as well as in the female hypothalamus and pituitary from
hatch to nestingly care day 5.

\hypertarget{fig-2-ability-to-use-gene-expression-to-distingush-groups}{%
\subsection{Fig 2: Ability to use gene expression to distingush
groups}\label{fig-2-ability-to-use-gene-expression-to-distingush-groups}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}

\hypertarget{fig-3-prolactin-expression-in-pit-but-not-hyp-or-gon-follows-same-pattern-as-in-blood}{%
\subsection{Fig 3: Prolactin expression in PIT (but not HYP or GON)
follows same pattern as in
blood}\label{fig-3-prolactin-expression-in-pit-but-not-hyp-or-gon-follows-same-pattern-as-in-blood}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}

\hypertarget{fig-4-hundreds-of-genes-show-similar-pattern-in-expression-as-circulating-prolactin}{%
\subsection{Fig 4: Hundreds of genes show similar pattern in expression
as circulating
prolactin}\label{fig-4-hundreds-of-genes-show-similar-pattern-in-expression-as-circulating-prolactin}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-6-1.pdf}

>>>>

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.

@@

Understanding the mechanisms underlying parental care are critical to
circumventing issues with parent-newborn bonding as well, where ultimate
explanations are obvious, but specific mechanisms remain elusive. The
rock dove (\emph{Columba livia}) is an ideal system to characterize
changes in genetic expression during parental care transitions because:
1) ample genomic resources are available (Gillespie et al.~2013),
including a complete annotated genome assembly {[}1{]} and methodology
concerning reproductive physiology and behavior (Dong et al.~2012); and
2) rock doves are prolific, year-round breeders that thrive in
captivity, making observation, manipulation, and sampling highly
feasible year-round. Rock doves are socially monogamous and offer
bi-parental care, making inter- and intra-sexual comparisons possible.
Birds offer two important behavioral transition points into parental
care: the incubation of eggs and the caring for chicks. This produces
two unique opportunities to study how the brain transitions into two
different suites of parental care behaviors. Additionally, rock doves
exhibit a parental care strategy analogous to mammals in that they, too,
`lactate' to feed their young (Gillespie et al.~2011, 2012). This
lactation, unlike simple regurgitation of food, consists of the
production and sloughing off of skin cells inside the crop sac of
females and males, creating a protein-rich milk-like substance on which
they rear their chicks. Many functional similarities between rock dove
and mammalian lactation exist concerning the mediation of this event by
the hormone prolactin (Dumont 1965). Additionally, like mammalian milk,
rock dove milk delivers essential immunoglobulins and nutritional
benefits to young, aiding in their immune function and development of
microbiota {[}2{]}. Thus, because rock doves incubate eggs and exhibit
mammalian-like mediation and function of lactation for young, they have
the potential to serve as a powerful theoretical bridge to understand
the neurobiology of both avian and mammalian transitions into parental
care. Our working hypothesis is that distinct changes in transcription
occur in the brain at the anticipation of, during, and in response to
two different types of parental care: incubation behavior and hatchling
care.

@@

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Experimental design.We sampled 9 timepoints, including 1
non-parental control, 1 nest-building, 1 egg lay, 3 incubation time
point, 1 hatch, and 2 nestling care stages. we collected truck blood,
the hypothalamus, the pituitary, and the gonads from all parents.}
\end{figure}

EdgeR and DESeq2 were used to calculate statistical differnces in gene
expression between each transition. All subsequent analyses were
conducting using variance stabilized data from the DESeq2 model. A
linear discriminant analysis was used to determine the confidence with
which we could predict the parental care stage of a sample based on it's
global gene expression profile. WGCNA was used to identify genes who's
expression was correlated with specific candidate genes of interest.

@@

Thosands of genes were found to be differentially expressed between the
non-parental control group and the nest-building group (Table 1, Fig.
1). Within the time course of parental care, fewer than 1000 genes
change there expression over the course of a few days, with the most
pronounced difference in toltal number of changes occuring in the female
pituitary between incubation days 9 and 17 (Table 1, Fig. 2).

\begin{table}[hb]
\begin{tabular}{llllllll}
\hline
\textbf{Method} & \textbf{Contrast} & \textbf{Hyp F} & \textbf{Hyp M} & \textbf{Pit F} & \textbf{Pit M} & \textbf{Gon F} & \textbf{Gon M} \\  \hline
EdgeR & Cont - Bldg & 14674 & 117 & 14593 & 144 & 14859 & 4 \\
EdgeR & Bldg - Lay & 14742 & 0 & 14593 & 0 & 14836 & 0 \\
EdgeR & Lay - Inc3 & 0 & 0 & 0 & 0 & 19 & 0 \\
EdgeR & Inc3 - Inc9 & 0 & 1 & 0 & 3 & 2 & 0 \\
EdgeR & Inc9 - Inc 17 & 0 & 2 & 57 & 37 & 9 & 0 \\
EdgeR & Inc17 - Hatch & 0 & 0 & 3 & 1 & 3 & 0 \\
EdgeR & Hatch - Nest5 & 1 & 0 & 3 & 0 & 0 & 0 \\
EdgeR & Nest5 - Nest9 & 0 & 0 & 0 & 0 & 24 & 0 \\  \hline
DESeq2 & Cont - Bldg & 3452 & 4499 & 3950 & 4680 & 4597 & 2135 \\
DESeq2 & Bldg - Lay & 0 & 0 & 101 & 0 & 104 & 1 \\
DESeq2 & Lay - Inc3 & 0 & 0 & 18 & 1 & 440 & 1 \\
DESeq2 & Inc3 - Inc9 & 1 & 0 & 0 & 1 & 8 & 0 \\
DESeq2 & Inc9 - Inc 17 & 1 & 0 & 640 & 186 & 1 & 0 \\
DESeq2 & Inc17 - Hatch & 2 & 0 & 3 & 18 & 1 & 0 \\
DESeq2 & Hatch - Nest5 & 166 & 0 & 340 & 3 & 1 & 0 \\
DESeq2 & Nest5 - Nest9 & 0 & 0 & 1 & 23 & 19 & 0 \\ \hline
\end{tabular}
\caption{Comparison of total DEGs determined by edgeR and DESeq2}
\label{tab:my-table}
\end{table}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{The magnitude of gene expression changes between each parental
transition. Using DESeq 2, we identified 4-5K genes are differentially
expressed beteen control birds and their nest buliding conspecifics in
all tissues except the male gonad. 500 - 1000 genes are differentially
expressed in the female pituitary from mid-late incubation as well as in
the female hypothalamus and pituitary from hatch to nestingly care day
5.}
\end{figure}

Even though we have a large sample size of roughly 12 samples per tissue
per sex, the differences between all 8 sampled parental time points are
so small, that we are only able to predict the stage using gene
expression along 40\%, 30\% and 27\% of the time in the pituiary, gonads
and hypothalamus, respectively (Fig 3A). However, if we group the
parental stages into anticipatory (control, nest building), egg care
(lay and incubation days), and nestling care (hatch and nesting days),
we incrase our ability to detect parental stage to 47\%, 43\%, and 37
pituiary, hypothalamus and gonads, respectively (Fig 3B). The

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Linear discriminant analysis shows the strongest ability to
predict parental stage of pitutary samples.}
\end{figure}

Gene expression of prlactin (\emph{PRL})in the pitutiary, but not the
hypothalamus follows the characteristic dip in expression relative to
non-parental controls during early incubation then spikes the day before
hatch and peaks at hatch (Fig 4). Intrestingly, hundreds of genes follow
this same, prolactin-associated pattern of expression in the pitutiary
(Fig. 5).

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Prolactin expression in PIT (but not HYP or GON) follows same
pattern as in blood}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Hundreds of genes show similar pattern in expression as
circulating prolactin}
\end{figure}
